---
title: "PS270 Final Project"
author: Kathy Bian
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---


## Project thoughts

I think I might do something with voting behavior among young people or Asian Americans. Maybe data about how often young people vote in certain elections in swing states vs non swing states. 


```{r}
library(readxl)
library(tidyverse)
```



```{r}
post <-
  read_excel("/Users/kbian50/PS 270/Final Project/data/Post-Dobbs State Abortion Restrictions and Protections_Data Page.xlsx")
```

```{r}
abortion_2022 <- 
  read_excel("/Users/kbian50/PS 270/Final Project/data/Abortions-Distributed-by-State-2013-2022.xlsx", sheet="2022", skip = 1)
```
```{r}
head(abortion_2022)
abortion_2022 <- abortion_2022 |>
  select(Area, `Total by location of service`)
```

```{r}
head(post)
```


Did states that enacted total abortion bans after Dobbs experience statistically significant declines in in-state abortion rates relative to states that did not? In this study, I want to examine whether or not abortion restrictions in 2022 actually decrease the rate of abortions performed in each state. I hypothesize that abortion rates decreased, but not in a drastic fashion that anti-abortion politicians (Republicans) intend. The Republican Party platform expects a direct causal relationship between abortion restrictions and rates of abortions-- if abortion is restricted or banned, abortion rates will directly go down. However, I don't believe this accounts for confounding variables that can affect abortion rates, such as interstate travel, tele-health, etc. I plan on focusing on a specific state or comparing one state that didn't enact a ban vs. a state that did not enact a ban. For example, the explanatory variable is a state that enacted a post-Dobbs ban and a state that did not. The outcome variable of interest is the rate of abortions performed. If there is a slight decrease in abortion rates, this will prove my hypothesis. If there is no change or an increase, this will disprove my hypothesis.



## Final Project Milestone 3 

```{r}
library(readxl)
library(tidyverse)
```



```{r}
post <-
  read_excel("/Users/kbian50/PS 270/Final Project/data/Post-Dobbs State Abortion Restrictions and Protections_Data Page.xlsx")
post
```

```{r}
abortion_2022 <- 
  read_excel("/Users/kbian50/PS 270/Final Project/data/Abortions-Distributed-by-State-2013-2022.xlsx", sheet="2022", skip = 1)
abortion_2022
```
```{r}
abortion_2022 <- abortion_2022 |>
  select(Area, `Total by location of service`)|>
  na.omit(abortion_2022) |>
     filter(`Total by location of service` != "--") |>
    rename(`Total number of abortions in the area` = `Total by location of service`) |>
    mutate(Year = 2022,
           `Total number of abortions in the area` = as.numeric(`Total number of abortions in the area`))
abortion_2022
```

```{r}
abortion_2021 <- 
  read_excel("/Users/kbian50/PS 270/Final Project/data/Abortions-Distributed-by-State-2013-2022.xlsx", sheet="2021", skip = 1)
abortion_2021
```

```{r}
abortion_2021 <- abortion_2021 |>
  select(Area, `Total by location of service`)|>
  na.omit(abortion_2021) |>
     filter(`Total by location of service` != "--") |>
    rename(`Total number of abortions in the area` = `Total by location of service`) |>
    mutate(Year = 2021,
            `Total number of abortions in the area` = as.numeric(`Total number of abortions in the area`))
abortion_2021
```

```{r}
abortion_combined <-
  rbind(abortion_2021, abortion_2022)
abortion_combined
```

```{r}
library(scales)
abortion_hist <-
  ggplot(data = abortion_combined,
         mapping = aes(x = `Total number of abortions in the area`)) +
  geom_histogram(bins = 20) + facet_wrap(~Year) + 
  scale_x_log10(labels = label_comma()) + 
  labs(x = "Number of Abortions",
       y = "Number of States",
       title = "Comparison of Post-Dobbs Abortion Rates (2021 vs 2022)")
abortion_hist
```




