---
title: "PS270 Final Project"
author: Kathy Bian
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---


## Project thoughts

I think I might do something with voting behavior among young people or Asian Americans. Maybe data about how often young people vote in certain elections in swing states vs non swing states. 


```{r}
library(readxl)
library(tidyverse)
```



```{r}
post <-
  read_excel("/Users/kbian50/PS 270/Final Project/data/Post-Dobbs State Abortion Restrictions and Protections_Data Page.xlsx")
```

```{r}
abortion_2022 <- 
  read_excel("/Users/kbian50/PS 270/Final Project/data/Abortions-Distributed-by-State-2013-2022.xlsx", sheet="2022", skip = 1)
```
```{r}
head(abortion_2022)
abortion_2022 <- abortion_2022 |>
  select(Area, `Total by location of service`)
```

```{r}
head(post)
```


Did states that enacted total abortion bans after Dobbs experience statistically significant declines in in-state abortion rates relative to states that did not? In this study, I want to examine whether or not abortion restrictions in 2022 actually decrease the rate of abortions performed in each state. I hypothesize that abortion rates decreased, but not in a drastic fashion that anti-abortion politicians (Republicans) intend. The Republican Party platform expects a direct causal relationship between abortion restrictions and rates of abortions-- if abortion is restricted or banned, abortion rates will directly go down. However, I don't believe this accounts for confounding variables that can affect abortion rates, such as interstate travel, tele-health, etc. I plan on focusing on a specific state or comparing one state that didn't enact a ban vs. a state that did not enact a ban. For example, the explanatory variable is a state that enacted a post-Dobbs ban and a state that did not. The outcome variable of interest is the rate of abortions performed. If there is a slight decrease in abortion rates, this will prove my hypothesis. If there is no change or an increase, this will disprove my hypothesis.



## Final Project Milestone 3 

```{r}
library(readxl)
library(tidyverse)
```



```{r}
post <-
  read_excel("/Users/kbian50/PS 270/Final Project/data/Post-Dobbs State Abortion Restrictions and Protections_Data Page.xlsx")
post
```

```{r}
abortion_2022 <- 
  read_excel("/Users/kbian50/PS 270/Final Project/data/Abortions-Distributed-by-State-2013-2022.xlsx", sheet="2022", skip = 1)
abortion_2022
```
```{r}
abortion_2022 <- abortion_2022 |>
  select(Area, `Total by location of service`)|>
  na.omit(abortion_2022) |>
     filter(`Total by location of service` != "--") |>
    rename(`Total number of abortions in the area` = `Total by location of service`) |>
    mutate(Year = 2022,
           `Total number of abortions in the area` = as.numeric(`Total number of abortions in the area`))
abortion_2022
```

```{r}
abortion_2021 <- 
  read_excel("/Users/kbian50/PS 270/Final Project/data/Abortions-Distributed-by-State-2013-2022.xlsx", sheet="2021", skip = 1)
abortion_2021
```

```{r}
abortion_2021 <- abortion_2021 |>
  select(Area, `Total by location of service`)|>
  na.omit(abortion_2021) |>
     filter(`Total by location of service` != "--") |>
    rename(`Total number of abortions in the area` = `Total by location of service`) |>
    mutate(Year = 2021,
            `Total number of abortions in the area` = as.numeric(`Total number of abortions in the area`))
abortion_2021
```

```{r}
abortion_combined <-
  rbind(abortion_2021, abortion_2022)
abortion_combined
```
```{r}
abortion_combined <- abortion_combined |>
  mutate(area = ifelse(Area %in% c("New York City", "New York State"),
                       "New York",
                       Area)) |>
  group_by(Area, Year) |>
  summarize(`Total number of abortions in the area` = sum(`Total number of abortions in the area`),
            .groups = "drop")
abortion_combined
```


```{r print-code, ref.label=knitr::all_labels(), echo = FALSE, eval=FALSE}
library(scales)
abortion_hist <-
  ggplot(data = abortion_combined,
         mapping = aes(x = `Total number of abortions in the area`)) +
  geom_histogram(bins = 20) + facet_wrap(~Year) + 
  scale_x_log10(labels = label_comma()) + 
  labs(x = "Number of Abortions",
       y = "Number of States",
       title = "Comparison of Post-Dobbs Abortion Rates (2021 vs 2022)") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
abortion_hist
```

## Final Project Milestone 4 

```{r}
post
```
```{r}
cleaned_post <- post |>
  group_by(Jurisdiction) |>
  summarise(legalstat = max(legalstat, na.rm = TRUE)) |>
  ungroup()
cleaned_post
```
```{r}
abortion_combined_fixed <- abortion_combined |>
  mutate(Area = if_else(Area %in% c("New York City", "New York State"),
                       "New York",
                       Area)) |>
  group_by(Area, Year) |>
  summarise(`Total number of abortions in the area` = 
              sum(`Total number of abortions in the area`),
            .groups = "drop")
abortion_combined_fixed
```
```{r}
abortion_combined_final <- abortion_combined_fixed |>
  left_join(cleaned_post, by = c("Area" = "Jurisdiction")) |>
  drop_na()
abortion_combined_final
```

```{r}
combined_boxplot <-
  ggplot(data = abortion_combined_final,
         mapping = aes(x = as.factor(legalstat), y = `Total number of abortions in the area`)) +
  geom_boxplot(fill = "steelblue", alpha = 0.8) +
  geom_jitter(width = 0.2, alpha = 0.5) + 
  labs(title = "Number of Abortions Performed in States after Post-Dobbs Restrictions",
       x = "Law Passed (O = No, 1 = Yes)",
       y = "Number of Abortions")
combined_boxplot
```
My research question asks whether post-Dobbs abortion restrictions passed in states affect the number of abortions in that particular state. This boxplot shows that states that did not pass post-Dobbs abortion restrictions have a higher median of abortions performed, compared to states that did pass post-Dobbs abortion restrictions. This is a logical relationship and to be expected. However, in states that did pass restrictions, there are more extreme outliers, with some even exceeding 80,000 abortions in that state. This can be represented by states with high populations, like California, or can demonstrate people from out-of-state arriving in states with no restrictions to get an abortion. 

```{r}
library(broom)
library(knitr)
library(kableExtra)
abortion.reg <- lm(`Total number of abortions in the area` ~ legalstat, data = abortion_combined_final)
abortion.reg
```
```{r}
tidy_model<- tidy(abortion.reg)
tidy_model

tidy_model |>
  kable(col.names = c("Term", "Estimate", "Std. Error", "Statistic", "p-value"), digits = 2, caption = "Regression Results: Impact of Post Dobbs Abortion Restrictions") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```

The average number of abortions in states that did not pass Post-Dobbs abortion laws is about 15,186. On average, states that passed restrictive laws have 2,851 fewer abortions than states that did not.


## HIGHLIGHT IMPORTANT INFO 
*italics* 
**bold** 

Clean up project milestones
Explain things to a person who hasn't taken this class 